---
title: "Advanced R Chapter 10"
subtitle: "Function Factories"
author: "Sam Bryce-Smith, R4DS Reading Group, Cohort 3"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    includes:
      after_body: insert-logo.html

---

```{r required packages, include=FALSE}
library(rlang)
library(ggplot2)
library(scales)
library(purrr)
```


```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_light(
  base_color = "#053F61",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Droid Mono"),
)
```

# What are function factories?

- Function factories are functions that produce functions
- *power1* - function factory
- *square* & *cube* - 'manufactured' function produced by function factory (and assigned to name)

```{r a basic function factory}

power1 <- function(exponent) {
  function(x) {
    x ^ exponent
  }
}

square <- power1(2)
cube <- power1(3)

square(2)
cube(2)
```

---

### How do they work? A recap on key concepts of functions

- How does R find the value of exponent?

- A new environment is created every time a function is called
  - Function is executed inside this environment (execution environment)
  - Binds/captures the enclosing environment when it is created (function environment)
    - (By default this is the current environment)
  
```{r}
y <- 1
f <- function(x) x + y
fn_env(f)
env_print(f)
#> <environment: R_GlobalEnv>
```

- Lexical scoping - heirarchy to finding names
  - Current function (exection environment)
  - Where function was defined (environment enclosing function, function environment)
  - Further levels up (all the way up to global environment)
  
  
---

# How do they work? A recap on key concepts of functions

```{r a basic function factory returns}

power1 <- function(exponent) {
  function(x) {
    x ^ exponent
  }
}

square <- power1(2)
cube <- power1(3)

square(2)
cube(2)
```

- power1(2) creates execution environment (where exponent = 2)
- function(x) is created, and binds the environment enclosing it (the execution environment of power1(2))


---
# Environments

square and cube have the same function body, but two different environments.

```{r square & cube envs}
square
cube
```

---

# square and cube have bindings to exponent

```{r}
env_print(square)

env_print(cube)

```

Square and cube have bindings to exponent - defined in the execution environment of power1 (the function factory)

---

# What are values of exponent in square and cube?

use *rlang::fn_env()* to extract the function environment

```{r accessing values in env}
fn_env(square)$exponent

fn_env(cube)$exponent
```

exponent is bound to different values in each case

---

# Be extra careful with function factories and lazy evaluation

**Lazy evaluation** - R only evaluates arguments when they are needed 

```{r}
x <- 2
square <- power1(x)
x <- 3

square(2)
```

- Value bound to x has changed between calling the *factory function* (to create the function) and calling the *manunfactured function* (when value of x is needed and looked up).
---

# Be sure to use (the) force()

If the argument is used by the manufactured function, you should force it's evaluation with **force()**

```{r good use of the force}

power2 <- function(exp) {
  force(exp)
  function(x) {
    x ^ exp
  }
}

x <- 2
square <- power2(x)
x <- 3
square(2)

```
---

# Stateful functions

- Function factories let you maintain state across invocations
  - Manufactured functions have their own, constant enclosing environment
  
`<-` - binds a value to a name in the **current environment**
`<<-` - *rebinds** an existing name found in a **parent environment**

```{r counter function}
new_counter <- function() {
  i <- 0
  
  function() {
    i <<- i + 1
    i
  }
}

counter_one <- new_counter()
counter_two <- new_counter()

counter_one()

counter_one()

counter_two()

```

- Manufactured functions have their own unique enclosing environment 
- counter_one and counter_two will have independent counnts
- Best used in moderation

---
# Garbage collection

As manufactured functions bind the execution environment of the function factory, temporary objects that are no longer needed won't be removed.

You should explicitly use **rm()** to unbind objects in function factories

```{r explicit garbage collection}
f1 <- function(n) {
  x <- runif(n)
  m <- mean(x)
  function() m
}

g1 <- f1(1e6)
lobstr::obj_size(g1)
#> 8,013,104 B

f2 <- function(n) {
  x <- runif(n)
  m <- mean(x)
  rm(x)
  function() m
}

g2 <- f2(1e6)
lobstr::obj_size(g2)
#> 12,944 B
```

# 10.2.6 Exercises

1. Why is it better to force(x) instead of just x?

```{r}
force
#> function (x) 
#> x
#> <bytecode: 0xce23a0>
#> <environment: namespace:base>
```

My answer - being explicit can help with clarity of code... 

```{r}

power2 <- function(exp) {
  force(exp)
  function(x) {
    x ^ exp
  }
}

x <- 2
square <- power2(x)
x <- 3
square(2)

power2_just_exp <- function(exp) {
  exp
  function(x) {
    x ^ exp
  }
}

z <- 2
square_just_exp <- power2_just_exp(z)
z <- 3
square_just_exp(2)

```

---
# 10.2.6 Exercises

2. What do approxfun() and ecdf() do?


---

---
# 10.2.6 Exercises

3. Create a function pick() that takes an index, i, as an argument and returns a function with an argument x that subsets x with i.

My answer -
want i to be input to function factory
Want to make sure i is evaluated when function is called
Want manufacured function to have binding to i

```{r solution to 10.2.6 Q3}

pick <- function(i) {
  force(i)
  
  function(x) {
    x[[i]]
  }
}


expected_q3 = lapply(mtcars, function(x) x[[5]])
identical(expected_q3,
          lapply(mtcars, pick(5)))
```

---

# 10.2.6 Exercises

4. Create a function that creates functions that compute the ith central moment of a numeric vector.



---

# 10.2.6 Exercises

5. What happens if you don’t use a closure? Make predictions, then verify with the code below.

The function will still count the number of invocations, but i isn't captured in enclosing environment of new_counter2 (it's in global environment)

I could modify i without having to call the function


```{r}

i <- 0
new_counter2 <- function() {
  i <<- i + 1
  i
}

new_counter2()
new_counter2()
i <- 0
new_counter2()

```

---

# 10.2.6 Exercises

6. What happens if you use <- instead of <<-? Make predictions, then verify with the code below.

Every time the function is called, i is updated in the execution environment only. So every time time the function is called again, i is still 0 (the enclosing environment for the manufactured function is constant unless bindings are updated with <<-)

This should return 1 every time

```{r}

new_counter3 <- function() {
  i <- 0
  function() {
    i <- i + 1
    i
  }
}
```

---
# Applications of function factories - ggplot2

ggplot2 & scales provide nice opportunity to use function factories

*scales* provides functions to control details of axes and legends
- the formatter functions are function factories 

```{r}

y <- c(12345, 123456, 1234567)
comma_format()(y)
#> [1] "12,345"    "123,456"   "1,234,567"

number_format(scale = 1e-3, suffix = " K")(y)
#> [1] "12 K"    "123 K"   "1 235 K"


```

---

# ggplot scales accept functions for the label argument


```{r}

df <- data.frame(x = 1, y = y)
core <- ggplot(df, aes(x, y)) + 
  geom_point() + 
  scale_x_continuous(breaks = 1, labels = NULL) +
  labs(x = NULL, y = NULL)
  
core
core + scale_y_continuous(
  labels = comma_format()
)
core + scale_y_continuous(
  labels = number_format(scale = 1e-3, suffix = " K")
)
core + scale_y_continuous(
  labels = scientific_format()
)


```

---

# Histograms - binwidth 
- *geom_histogram(binwidth)* can accpet a function as the binwidth argument
- This allows you to have **different binwidths for plots in different facets**

```{r}
# construct some sample data with very different numbers in each cell
sd <- c(1, 5, 15)
n <- 100

df <- data.frame(x = rnorm(3 * n, sd = sd), sd = rep(sd, n))

ggplot(df, aes(x)) + 
  geom_histogram(binwidth = 2) + 
  facet_wrap(~ sd, scales = "free_x") + 
  labs(x = NULL)
```


Given the wide variability in counts between 1, 5 & 15, it is less clear to see the counts in 5 & 15...

---

# Histograms - binwidth

Use a function factory to get approximately the same number of observations in each bin
- Input desired number of bins
- Output function taking a numeric vector that returns a binwidth

```{r}

binwidth_bins <- function(n) {
  force(n)
  
  function(x) {
    (max(x) - min(x)) / n
  }
}

```

---

# Histograms - binwidth

```{r}
ggplot(df, aes(x)) + 
  geom_histogram(binwidth = binwidth_bins(20)) + 
  facet_wrap(~ sd, scales = "free_x") + 
  labs(x = NULL)

```

---

# Histograms - binwidth

- `class.Sturges()`, `nclass.scott()` and `nclass.FD()` are base functions to find optimal binwidth
- Can wrap around these with a function factory

```{r}
base_bins <- function(type) {
  fun <- switch(type,
    Sturges = nclass.Sturges,
    scott = nclass.scott,
    FD = nclass.FD,
    stop("Unknown type", call. = FALSE)
  )
  
  function(x) {
    (max(x) - min(x)) / fun(x)
  }
}

```

---

# Histograms - binwidth

```{r}
ggplot(df, aes(x)) + 
  geom_histogram(binwidth = base_bins("FD")) + 
  facet_wrap(~ sd, scales = "free_x") + 
  labs(x = NULL)
```

---

# Function factories for statistics

- Further examples for using function factories
- The Box-Cox transformation
- Bootstrap resampling
- Maximum likelihood estimation

(Disclaimer - I only dove into bootstrap resampling, but I think it demonstrates previous concepts nicely)

---

# Bootstrapping

- resampling method
- randomly sample with replacement (& repeat numerous times)
- Can be used to estimate statistics about a population

---

# Bootstrapping

```{r}

boot_permute <- function(df, var) {
  n <- nrow(df)
  force(var)
  
  function() {
    col <- df[[var]]
    col[sample(n, replace = TRUE)]
  }
}

boot_mtcars1 <- boot_permute(mtcars, "mpg")
head(boot_mtcars1())
#> [1] 16.4 22.8 22.8 22.8 16.4 19.2
head(boot_mtcars1())
#> [1] 17.8 18.7 30.4 30.4 16.4 21.0

```

---

## Useful for parametric bootstrap
- Fit model once, then explicitly remove temporary object (linear model objects are large)

```{r}
boot_model <- function(df, formula) {
  mod <- lm(formula, data = df)
  fitted <- unname(fitted(mod))
  resid <- unname(resid(mod))
  rm(mod)

  function() {
    fitted + sample(resid)
  }
} 

boot_mtcars2 <- boot_model(mtcars, mpg ~ wt)
head(boot_mtcars2())

head(boot_mtcars2())


```

---

# 10.4.4 Exercises

I haven't tackled any of these - sorry!

---

### Combining function factories with functionals

- Create many power functions by iterating over named list

```{r}
names <- list(
  square = 2, 
  cube = 3, 
  root = 1/2, 
  cuberoot = 1/3, 
  reciprocal = -1
)
funs <- purrr::map(names, power1)

funs$root(64)
funs$root
```

- Functions are stored in list *funs*
- calling function with prefix *funs$* could be annoying...

---
### How to avoid prefixing function call with funs$

- Can use *with()*

```{r}
with(funs, root(100))
```
- Only temporary

- *attach()* & *detach()* is longer lasting approach (but be careful)
```{r}
attach(funs)
root(100)
detach(funs)
```

---

### How to avoid prefixing function call with funs$

- Copy functions to global environment with *rlang::env_bind()*
- (Pretty much permanent)

```{r}
rlang::env_bind(globalenv(), !!!funs)
root(100)
```

- Can be unbound later (*rlang::env_unbind()*), but names could have been rebound in intervening period
  - Could end up deleting object not created by binding funs
  
```{r}
rlang::env_unbind(globalenv(), names(funs))
```

- more about quasiquotation - `!!!` - in **Section 19**...

---

# 10.5.1 - Exercises

Q - **Which of the following commands is equivalent to with(x, f(z))?**

- x$f(x$z).
- f(x$z).
- x$f(z).
- f(z).
- It depends.


My initial thoughts...

- If x is a list of functions as in previous example, then
  - `x$f(z)` would be my answer 


---

# 10.5.1 - Exercises

A - Depends on structure of `x`!
- if x is a dataframe, then `f(x$z)`
- if x is a list, then it could be any of the options!
    
```{r}
f <- mean
z <- 1
x <- list(f = mean, z = 1)

identical(with(x, f(z)), x$f(x$z))

identical(with(x, f(z)), f(x$z))

identical(with(x, f(z)), x$f(z))

identical(with(x, f(z)), f(z))
```

---

class: inverse, hide-logo

# Another Slide
This slide doesn't have a logo

---

